Columnar VCF for Biobank scale datasets using Zarr

### 2024-04-26 paper.tex
Some work on the intro narrative

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale projects such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
While there are many advantages to encoding data in text files,
VCF is simply not fit for purpose as part of computational workflows 
at this scale. [A sentence somehow establishing that methods analysing 
large-scale genomic data are held back by VCF, needing to first 
convert to intermediate formats like tsinfer.]

@@

VCF: Variant Call Format;

\subsection{Competing Interests}

\subsection{Funding}

>>>>

\section{Background}
Variant Call Format (VCF) is the standard format for interchanging genetic
variation data~\citep{danecek2011variant}, storing information about 
DNA sequence polymorphisms among a set of samples with associated 
quality control metrics and metadata. 
Originally defined specifically as a text file,
it has been refined and standardised~\citep{rehm2021ga4gh} and the 
underlying data-model is now deeply embedded in bioinformatics practice.
Dataset sizes have grown explosively since the introduction of 
VCF as part of 1000 Genomes project~\citep{10002015global},
with Biobank-scale projects such as 
Genomics England~\cite{turnbull2018100},
UK Biobank~\citep{bycroft2018genome,backman2021exome,halldorsson2022sequences,uk2023whole},
and the All of Us research program~\citep{all2024genomic}
collecting genome sequence data for hundreds of thousands of humans.
While there are many advantages to encoding data in text files,
VCF is simply not fit for purpose as part of computational workflows 
at this scale. 
Methods that require efficient access to data
either require VCF data to be converted into 
PLINK~\cite{purcell2007plink,chang2015second} 
or BGEN~\citep{band2018bgen} 
format~\citep[e.g.][]{yang2011gcta,mbatchou2021computationally,loh2015efficient}
or use 
custom binary formats~\citep[e.g.][]{
% Uses custom "bref3" format,
% https://faculty.washington.edu/browning/beagle/bref3.24May18.pdf
browning2018one, 
% .samples Zarr format
kelleher2019inferring,
% Has a "xcftools" package, but it still looks pretty experimental
hofmeister2023accurate} 
that support the required access patterns.
While PLINK and BGEN support efficient access to genotype data, neither
can accommodate the full flexibility of the VCF data model and conversion
is lossy. PLINK's approach of storing the genotype matrix in uncompressed
packed-binary format also leads to large files, particularly for recent
whole-genome sequence datasets. For example, at two bits 
per genotype, the full matrix for the GraphTyper SNP dataset in the UKB
WGS data~\citep{uk2023whole} is 116 TiB.
% 1,037,556,156 SNPs x  490,640 samples
% humanize.naturalsize(1_037_556_156 * 490_640 / 4, binary=True)
% '115.7 TiB'

@@

\begin{itemize}
    \item VCF: Variant Call Format;
    \item UKB: UK Biobank
    \item WGS: Whole Genome Sequence
\end{itemize}

\subsection{Competing Interests}
\subsection{Funding}
